---
title: "Single cell analysis"
author:
  - "Valentin Gabeff"
  - "Théo Imler"
  - "Abigail Strefeler"
  - "léo Sumi"
date: "March 2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..")
```

```{r libraries, include=FALSE}
# Loading libraries
library(ggplot2)
library(ellipse)
library(matrixStats)
library(limma)
library(ggfortify)
library(pca3d)
library(factoextra)
library(pheatmap)
library(cluster)
library(rgl)
```

```{r import, include=FALSE}
# Load the data
data <- read.table("data/Group5_data.txt", header = TRUE, sep = "\t", dec = ".", row.names = "X")
data.cells <- ncol(data)
data.rows <- nrow(data)
```

## Introduction

A heterogenous mix of cells was analysed using single-cell RNA-seq. After pre-processing and quality control, we have a dataset containing 73 cells showing expression of 23'351 genes.
The number of read genes per cell, on average, comes out to 9503. This was calculated by simply counting the number of nonzero entries for each cell, and performing an average. The number of genes per cell can give information about quality, since samples epxressing too few genes are more likely free RNA, and samples expressing too many are likely doublets of cells. Plotting expression per cell shows an even distribution of gene counts over the different cells (TODO).

```{r genes per cell, echoe=FALSE}
# Genes per cell
per_cell <- colSums(data != 0)
avg_per_cell <- mean(per_cell)
#counts <- data.frame(per_cell, row.names=1:73)
#plot(counts)
hist(per_cell)
```

## Results

### Part 1: Number of cell types

The number of cell types is determined using clustering. The criteria we use for clustering is the expression of certain genes: cells that express these genes in a similar manner are more closely related than those that express them differently.

First we must decide which genes we will use to compare cells. We have two main options, either we retreive only highly expressed genes or we use genes with highly variable expression. Here we decided to use genes with a high variability as the histogram of variability shows an evident tail. Yet, this choice should not alter our results much. Hence out of all of the detected genes, we keep only these ones in our analysis. Explain why variablitlity is better as we use clusters

```{r marker genes, echo=FALSE}
# Get most expressed genes
gene.expression <- apply(data, 1, median)
hist(log10(gene.expression), breaks = 50, col='blue', main = "Gene expression distribution", xlab = "Gene expression median (log10)")

# Get Most variable genes
gene.variation <- apply(data, 1, sd)
hist(log10(gene.variation), breaks = 50,col = "blue",freq = TRUE,main = "Gene variation distribution", xlab = "Gene expression standard deviation (log10)") 
abline(v = 3.61, col = "red",lwd=3, lty=2)

# Because we choose to use the 10% most variable genes, the most expressed genes object is no more necessary
rm(gene.expression)

# We choose to recover the 10% most variable genes
sorted_sd <- sort(gene.variation, decreasing = TRUE)
genes.mostvariable <- head(sorted_sd, 0.1*data.rows)
hist(log10(genes.mostvariable), breaks = 50)
```

To be able to correctly compare expression between cell types, gene expression must be normalized. This is due to the difference in the number of reads per cells.

```{r normalization, echo=FALSE}
# Filter and normalize the data for the 10% most variable genes
data.filtered_variable <- data[c(names(genes.mostvariable)),]
data.norm = voom(data.filtered_variable, plot=TRUE, normalize.method = "quantile")
#head(data.norm$E, 5)
```

```{r saving data, include=FALSE}
# Save to txt format the normalized data (data.norm$E)
write.table(data.norm$E, file = "results/normalized_data.txt", append = FALSE, sep = "\t",
                        eol = "\n", na = "NA", dec = ".", row.names = TRUE,
                        col.names = TRUE, qmethod = c("escape", "double"),
                        fileEncoding = "utf8")

# Save to binary format the normalized data (data.norm$E)
save(data.norm, file = "results/datanorm.rdata")
```

So far, data are bein spread on 2335 dimensions. We use Principal Component Analysis (PCA) to project the data on a 3D space whose basis explain most of the variance among the data.

Reducing dimensionality allows us to visualize the data, and gives us some idea about what clusters may be present. `fviz_eig` function allows us to know the percentage of explained variance for each principal component. We choose to use three principal components and represent the principal component analysis (PCA) in 3D. On a 3D interactive plot, we can visually identify 4 clusters. 

To cluster the data, we use a more robust version of k-means called partitioning around medoids (PAM). Representative objects called medoids are found, and groups are creating by assigning each point to the nearest medoid. This method minimizes the sum of dissimilarities of observations to their closest representative object.

```{r pca and clustering, echo=FALSE}
# PCA
# Let's find out what are the three first principal components that explain
# most of the variance between the cells. Reduction of dimension from 2335 to 3.
pca.fit <- prcomp(t(data.norm$E), scale. = TRUE)
fviz_eig(pca.fit, main = "Variance explained by top PCs")

# PCA in a 2D plot
# autoplot(pca.fit)

# Saving first three PCs in a file
write.table(pca.fit$rotation[,1:3], file = "results/principal_components.txt")

# PAM clustering method (more robust than k-means)
# 3D plot

wss <- data.frame(matrix(ncol=1,nrow=15))
  for (i in 1:15) wss[i,1] <- pam(x=pca.fit$x[,1:3], k=i, metric="euclidean", stand=FALSE)$objective['swap']
plot(x = row.names(wss),y = wss$matrix.ncol...1..nrow...15., type = "o", xlab="Number of Clusters",
     ylab="Objective function minimum",pch = 15) + abline(v = 4,col = "red",lwd=3, lty=2, main = "Optimal number of clusters for k-medoid")

pca.pam <- pam(x=pca.fit$x[,1:3], k=4, metric="euclidean", stand=FALSE)
pca3d(pca=pca.fit$x, group=pca.pam$clustering, show.ellipses=TRUE, show.plane=TRUE, fancy=FALSE, new = FALSE, bg = "lightgrey", show.axes = TRUE, axes.color = "black")
#snapshotPCA3d(file="figures/clusters.png")

# You can again save the clusters in a file
write.table(pca.pam$clustering, file = "results/kmeans_clusters.txt")
```

### Part 2: Marker genes for different cell types

Different cell types will have certain genes that they express in a manner different from other cell types. These are called differentially expressed genes.
By choosing one cluster of cell and comparing the expression of their genes against the other clusters, one can statistically determine the fold change 
of expression using the limma package.

The 100 differentially expressed genes can be represented using a heatmap. We indeed see that the 4 different cell types show differential expression of these. When the most expressed marker genes are plotted onto the PCA, they clearly belong to the different cell types we have identified. [Code/gif required]

We can see that the first two clusters on the heat map are quite close to each others. This is also seen on the plot of the DE genes for each cluster where RGS4 is expressed mainly in one cluster, but also in another. This means that the segregation between these two groups of cell is less important than between the other clusters.

```{r DE genes, include=FALSE}
# DE Genes

#Find genes most important to explain first PC
head(sort(pca.fit$rotation[,1], decreasing = TRUE))

#Find genes that explain the most first cluster
fit.C1 <- lmFit(data.norm$E, model.matrix(~(pca.pam$clustering == 1)))
fit.C1 <- eBayes(fit.C1)

DE.C1 = topTable(fit.C1,n=Inf,adjust="fdr", sort.by = "logFC")
DE.genesC1<-row.names(head(DE.C1[(abs(DE.C1[,1])>1) & (DE.C1[,5]<0.05),],100))
write.table(DE.genesC1, file = "results/DEGenesC1.txt", append = FALSE, sep = "\n",
            eol = "\n", na = "NA", dec = ".", row.names = FALSE,
            col.names = FALSE, qmethod = c("escape", "double"),
            fileEncoding = "utf8", quote=FALSE)

#Find genes that explain the most second cluster
fit.C2 <- lmFit(data.norm$E, model.matrix(~(pca.pam$clustering == 2)))
fit.C2 <- eBayes(fit.C2)

DE.C2 = topTable(fit.C2,n=Inf,adjust="fdr", sort.by = "logFC")
DE.genesC2<-row.names(head(DE.C2[(abs(DE.C2[,1])>1) & (DE.C2[,5]<0.05),], 100))
write.table(DE.genesC2, file = "results/DEGenesC2.txt", append = FALSE, sep = "\n",
            eol = "\n", na = "NA", dec = ".", row.names = FALSE,
            col.names = FALSE, qmethod = c("escape", "double"),
            fileEncoding = "utf8", quote=FALSE)

#Find genes that explain the most third cluster
fit.C3 <- lmFit(data.norm$E, model.matrix(~(pca.pam$clustering == 3)))
fit.C3 <- eBayes(fit.C3)

DE.C3 = topTable(fit.C3,n=Inf,adjust="fdr", sort.by = "logFC")
DE.genesC3<-row.names(head(DE.C3[(abs(DE.C3[,1])>1) & (DE.C3[,5]<0.05),],100))
write.table(DE.genesC3, file = "results/DEGenesC3.txt", append = FALSE, sep = "\n",
            eol = "\n", na = "NA", dec = ".", row.names = FALSE,
            col.names = FALSE, qmethod = c("escape", "double"),
            fileEncoding = "utf8", quote=FALSE)

#Find genes that explain the most fourth cluster
design.C4 <- matrix(0, nrow = 73, ncol = 1)
design.C4[pca.pam$clustering == 4] = 1
colnames(design.C4) <- c("Cluster_4")

fit.C4 <- lmFit(data.norm$E, model.matrix(~(pca.pam$clustering == 4)))
fit.C4 <- eBayes(fit.C4)

DE.C4 = topTable(fit.C4,n=Inf,adjust="fdr", sort.by = "logFC")
DE.genesC4<-row.names(head(DE.C4[(abs(DE.C4[,1])>2) & (DE.C4[,5]<0.05),], 100))
#DE.genesC4<-row.names(DE.C4)[DE.C4$logFC>2 & DE.C4$adj.P.Val<0.05]
write.table(DE.genesC4, file = "results/DEGenesC4.txt", append = FALSE, sep = "\n",
            eol = "\n", na = "NA", dec = ".", row.names = FALSE,
            col.names = FALSE, qmethod = c("escape", "double"),
            fileEncoding = "utf8", quote=FALSE)
```

```{r heatmap, echo=FALSE}
pheatmap(data.norm[c(DE.genesC1, DE.genesC2, DE.genesC3, DE.genesC4),],
         cluster_cols = T,
         cluster_rows = F,
         show_colnames = F,
         show_rownames = F,
         annotation_col = data.frame(cluster = factor(pca.pam$clustering, labels = paste("Cluster", 1:4))),
         main = "Heatmap of the top 100 most expressed genes for each clusters"
         )
```

## Determining the cell types

To determine the different cell types present in our sample, we upload the differentially expressed genes for each cluster into EnrichR. We found that we have cells from the heart, adipose tissue, hypothalamus, and cerebral cortex. These last two categories explain the similarity between two of our clusters, since they are both neural tissues.

```{r cell type, echo= FALSE}
c4.expr = data.filtered_variable[DE.genesC4[1],]
c4.expr_norm = log10(c4.expr+1)/max(log10(c4.expr+1))
c4.expr_col = colorRamp(c("white", "red"))(c4.expr_norm)
pca3d(pca = pca.fit$x, col = rgb(c4.expr_col, maxColorValue = 255))

c1.expr = data.filtered_variable[DE.genesC1[1],]
c1.expr_norm = log10(c1.expr+1)/max(log10(c1.expr+1))
c1.expr_col = colorRamp(c("white", "red"))(c1.expr_norm)
max(c1.expr_norm)
min(c1.expr_norm)
pca3d(pca = pca.fit$x, col = rgb(c1.expr_col, maxColorValue = 255))

c2.expr = data.filtered_variable[DE.genesC2[1],]
c2.expr_norm = log10(c2.expr+1)/max(log10(c2.expr+1))
c2.expr_col = colorRamp(c("white", "red"))(c2.expr_norm)
max(c2.expr_norm)
min(c2.expr_norm)
pca3d(pca = pca.fit$x, col = rgb(c2.expr_col, maxColorValue = 255))

c3.expr = data.filtered_variable[DE.genesC3[1],]
c3.expr_norm = log10(c3.expr+1)/max(log10(c3.expr+1))
c3.expr_col = colorRamp(c("white", "red"))(c3.expr_norm)
max(c3.expr_norm)
min(c3.expr_norm)
pca3d(pca = pca.fit$x, col = rgb(c3.expr_col, maxColorValue = 255), title = DE.genesC3[1])
```

[Insert the graphs from EnrichR?]

[Litterature analysis: main marker genes that we observed have previously been found to be markers of heart, adipose tissue, etc.]
